Declarative Integration of Interactive 3D Graphics into the World-Wide Web:\\Principles, Current Approaches, and Research Agenda \\~

### 2013-03-09 Web3D2013-Dec3D/paper.tex
finished 1st version

Moreover, embedded 3D graphics should reuse existing W3C techniques (specifically from HTML5 and SVG) as far as possible and propose extensions only where 3D-specific features are necessary or where they provide significant benefits. Where new concepts are introduced their relation to and effects on existing Web standards should be analyzed, evaluated, and discussed with the respective W3C working groups.

@@

\paragraph{Platform Independence}
Another aspect is to describe 3D content in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only (cp. \cite{Schwenk12}). The technology investigated should allow for content to be portable across user agents, rendering techniques, and hardware platforms, while taking advantage of available features wherever possible. The results of rendering content under such different environments should be highly predictable.

@@

With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two experimental declarative 3D Web publishing frameworks are available. In the following we briefly introduce these open-source frameworks, which support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like. This discussion is followed by the presentation of the so-called ``polyfill'' approach for the further X3DOM and XML3D development activities.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.9\columnwidth]{images/polyfill.png}
  \caption{Levels of integration: current browser APIs only allow integrating with DOM and DOM Events, but not yet fully with CSS.}
  \label{fig:polyInt}
  \vspace{-0.2cm}
\end{figure}

% TODO: don't cite this link but present the whole idea here!!!
% http://www.w3.org/community/declarative3d/wiki/Level_of_Integration/
Recently, the group proposed to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS. etc.). The proposal aims to explain existing and possible integration levels (Figure \ref{fig:polyInt}) and finally to identify an integration level the group is aiming for.
% again, this is still bla bla
% instead, insert text from slides - see slide 22 ff.
It quickly became clear that a higher level of integration requires additional APIs in user agents. After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a polyfill approach for the further development activities.
Nevertheless, communicating and requesting additional and missing features like custom CSS properties or CSS monitoring is necessary.

@@

TODO % see slides, where these are discussed starting with slide 11/12

Dec3D Scene Graph
Modern and minimalistic scene graph or tree
traversal and culling
creates mesh/material/transform bundles for render layer
picking
Supports graph and tree structures for data reuse
API layer, no DOM integration!

Essential Dec3D DOM Elements
Focus on minimalistic element set, not usability
Represents SG concepts as DOM elements
Adds DOM, Events and CSS to SG

Comprehensive Dec3D DOM Elements
Builds on Essential and/or SG layer
Focus on usability not minimalism
Could build on existing SG standards (e.g. X3D) for convenience

1. Start element
Transition from HTML flow layout to 3D transform

2. CSS integration
Use exiting CSS properties where applicable
Define a set of 3D-specific CSS properties
Needs to be supported in UAs!

3. Hierarchy
HTML hierarchy
Comes with issues (e.g. CSS inheritance)

4. Transformations
Use CSS 3D Transforms

5. Generic Data Containers
Take modern HW into account
Name: Type, TupleSize, Value
Based on TypedArrays with interface to modify data

6. External Data Containers
External (binary) blobs / transmission format
Maps to generic data containers

7. Operators on Data Containers
Allows dynamic meshes, image processing, compression
Data driven approach: Xflow, X3D Routes, SVG Filter FX

8. Shaders
2-3 powerful fixed-function shaders
Enough for many use-cases
No timing attack issues
Can adapt to target device

9. Textures
Use HTML elements: \verb|<img>|, \verb|<video>|, \verb|<canvas>|
Interactive textures: \verb|<svg>|, \verb|<html>|

10. 3D Geometry?
Close to OpenGL primitives

11. Lights
Shader parameters in GL, but don't obstruct future concepts
Intuitive way to define lights

12. Viewpoints?
Link to viewpoints (intra and inter-document)

13. Camera navigation
Pre-defined navigation modes?

14. DOM Events
Support standard events: mouseover, click
3D-specific DOM Events contain 3D-specific context information

15. Interfaces and Types?
Provide IDLs for new elements
Convenience functions
Dec3D Base Types: Lean set of complex data types for 2D/3D
Provide / use types that help to manage 3D scene, e.g. CSS Matrix

@@

In this section we shape the agenda and identify upcoming research issues for the next development stage of Dec3D.
During the 1\textsuperscript{st} International Workshop on Declarative 3D for the Web Architecture \cite{Dec3D2012} as well as during more informal group meetings at Web3D and SIGGRAPH in 2011/2012, the members and supporters of the Declarative 3D W3C Community Group agreed that what is needed to make the effort successful and for the W3C to adopt/develop a Dec3D standard include the following topics.

>>>>

Embedded 3D graphics should reuse existing W3C techniques, specifically from HTML5 and SVG, as far as possible and propose extensions only where specific features are necessary or provide significant benefits. Where new concepts are introduced their relation to and effects on existing Web standards should be analyzed, evaluated, and discussed with the respective W3C working groups.

@@

\paragraph{Platform Independence}
3D content needs to be described in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only \cite{Schwenk12}. This should allow for content to be portable across browsers, rendering techniques, and hardware platforms, while taking advantage of available features wherever possible. The results of rendering content under such different environments should be highly predictable.

@@

With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two open-source frameworks are available, which are briefly introduced in this section. Both frameworks support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like. This discussion is followed by the presentation of the so-called ``polyfill'' approach for further development activities.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.85\columnwidth]{images/polyfill.png}
  \caption{Levels of integration: current browser APIs only allow integrating with DOM and DOM Events, but not yet fully with CSS.}
  \label{fig:polyInt}
  \vspace{-0.2cm}
\end{figure}

% Don't cite this link but present the whole idea here!
% http://www.w3.org/community/declarative3d/wiki/Level_of_Integration/
We also propose to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS, etc.) to explain existing and possible integration levels (Figure \ref{fig:polyInt} shows that currently only level 2, DOM integration, is possible), and to finally identify a suitable integration level for declarative 3D. However, the latter imposes several problems in that it requires having a set of 3D-specific CSS properties or even a full shader description via CSS as well as a deeper integration into debugging tools like Chrome Developer Tools or Firebug. However, integration levels 3 and 4 are currently not yet achievable by a JavaScript implementation, because the required APIs are missing: it is not possible to introduce new CSS properties or to extend developer tools by an API.

It quickly became clear that a higher level of integration requires additional APIs in user agents. After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a polyfill approach for the further development activities.
Nevertheless, communicating and requesting additional and missing features like custom CSS properties or CSS monitoring is necessary.

@@

In this section we briefly outline our proposed essential elements for an upcoming Dec3D standard. First of all, a start element allows the transition from the HTML flow layout to 3D transformations:

\small
\begin{verbatim}
<div>
    <dec3d style="border: 1px solid black;">
    ...
    </dec3d>
</div>
\end{verbatim}
\normalsize

The DOM integration obviously extends the HTML hierarchy. But this comes along with certain issues like the correct treatment of CSS inheritance. Therefore, CSS integration is another point in that existing CSS properties should be used where applicable. Transformations for instance can be represented using CSS 3D Transforms \cite{webkit3DCSS}. In addition, a set of 3D-specific CSS properties (e.g. for defining the appearance) needs to be defined. However, custom CSS properties are not yet supported by any Web Browser, though finally this seems to change now \cite{CSSCustom}.

Moreover, not only standard DOM UI Events like `mouseover' and `click' shall be supported, but also events that contain 3D-specific context information.
In this regard, the Browser API should expose convenience functions and special Dec3D base types consisting of a lean set of complex data types for 2D/3D graphics. Similarly, generic mesh data containers should be based on TypedArrays \cite{TypArr12} with appropriate interfaces to efficiently modify the data, where external binary containers are used as transmission format \cite{Behr12,Trevett12}. Thus, 3D geometry definition will stay close to OpenGL primitives.

Besides this, we further distinguish between essential Dec3D DOM Elements, were the focus lies on a minimalistic element set, as well as more comprehensive Dec3D DOM Elements (see middle of Figure \ref{fig:polyArch}), which build on top of those with the focus on usability and which for convenience could build on existing scene-graph standards like X3D \cite{Web3D-X3D}.

Essential DOM elements are a basic set of built-in shaders, which should be enough for most use cases. Advantages are that they do not entail timing attack issues and that they can adapt to any target device. As textures, HTML elements like \verb|<img>|, \verb|<video>|, \verb|<canvas>| shall be used, whereas elements like \verb|<svg>| and \verb|<html>| can serve as interactive textures.

Lights are basically also shader parameters in OpenGL/WebGL \cite{WebGL12}, but here shall be represented as own elements to provide intuitive ways to define lights and to not obstruct more advanced future concepts.
Likewise, viewpoints can serve as link to certain points of interest within the document. However, it is still discussable if some pre-defined camera navigation modes should be part of the essential profile.

@@

In this section we shape the agenda and identify upcoming research issues for the next development stage of Dec3D.
During the 1\textsuperscript{st} Intl. Workshop on Declarative 3D \cite{Dec3D2012} as well as during more informal meetings at Web3D and SIGGRAPH in 2011/2012, the members and supporters of the Declarative 3D W3C Community Group agreed upon the following topics to make the effort successful and for the W3C to adopt/develop a Dec3D standard.
